---
title: "session 8 dplyr and tidyr"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```


```{r}

library(dplyr)
library(tidyr)
library(readr)

```


## source data

### read in data

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```


## clean and reshape data


### remove excess columns

2 ways to do the same thing w/ select

```{r}

catch_data <- catch_original %>% select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

catch_data <- catch_original %>% select(-All, -notesRegCode)

head(catch_data)
summary(catch_data)

```

turn Chinook into numeric

```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))



which(is.na(catch_clean$Chinook)) #checks which one is an NA, i.e., not numeric

```



### check column typing
```{r}

#change the messed up "I" to a 1, then change it to numeric again since the if else thing turns 1 into character
#this is a good way to fix errors and stuff

catch_clean <-  catch_data %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook))

catch_original[401,]
catch_clean[401,]
```






### reshape data

make long and tidy
```{r}
# don't pivot region and year, but pivot everything else
catch_long <- catch_clean %>% 
  pivot_longer(-c(Region, Year), names_to = "species", values_to = "catch")
```

back to wide for fun
```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = "species", values_from="catch")

```

change column name to match fact that units are in 1000s of fish
```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands=catch)
```

make new column in for total number fish then get rid of the other one
```{r}
catch_long <- catch_long %>% 
  mutate(catch=catch_thousands*1000) %>% 
  select(-catch_thousands)

head(catch_long)
```


take mean per region (ignore the warning) then sort by highest to lowest catch
```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch)) %>% 
  arrange(desc(catch_mean))

mean_region

```

summarize count (number observations) by region
```{r}
n_region <- catch_long %>% 
  group_by(Region) %>%
  summarize(n=n())
```


filter for just SSE and check that that's the only one there
```{r}
SSE_catch <- catch_long %>% 
  filter(Region=="SSE")

unique(SSE_catch$Region)
```


## Join datasets (region and catch)

can read and modify/select at the same time
```{r}
region_defs <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
  select(code, mgmtArea)
```


join using the Region/code as foreign keys and also telling R that they are the same thing but with different names
```{r}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))
```


## separate and ünite

separate stuff from one column into 2 columns by the "-
```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAI-039"))

sites_df

sites_df %>% 
  separate(site, c("island", "site_number"), "-")
```

unite example
```{r}
dates_df <- data.frame(year = c("1930",
                                "1930",
                                "1930"),
                       month = c("12",
                                "12",
                                "12"),
                       day = c("14",
                               "15",
                               "16"),
                       stringsAsFactors = FALSE)

dates_df

dates_df %>% 
  unite(date, year, month, day, sep = "bawitdaba")
```

```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"),
                  stringsAsFactors = FALSE)
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1", method = "libcurl"),
                        stringsAsFactors = FALSE) %>% 
    select(code, mgmtArea)

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), names_to = "species", values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  left_join(region_defs, by = c("Region" = "code"))

```


